The need for accuracy and smoothness in numerical simulations

### 2024-05-14 report/src/draft.tex
Fixed typos

Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e, $\tau = 10^{-4}$. The amplitude of the oscillations is reduced when $\tau = 10^{-12}$. We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expectation for the computed approximation unless $\tau$ is very small.

@@

Thirdly, if the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} well inside the asymptotic range. Why is this? If we where in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and $p>0$. In particular, the value of $\hat{A}_h$ should behave in a \emph{monotone} manner and the tiny oscillations that we have recorded should not be present.

@@

When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, the experiments supports the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{iontrap_mwe2.pdf}
  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p$. The fields are zero outside a small sphere and discontinuous.}
  \label{fig:iontrap_mwe2}
\end{figure}
When the force-fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to support the existence of an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{iontrap_mwe4.pdf}
  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p$. The fields are zero outside a small sphere, but smooth.}
  \label{fig:iontrap_mwe4}
\end{figure}
When the force-fields vanish outside a small ball but are smooth, see Figure~\ref{fig:iontrap_mwe4}, we retain the ability to judge when Richardson's error estimates are reliable and we can quantify the effect of changing the force-field.

>>>>

Firstly, the potential energy and especially the kinetic energy exhibits violent oscillations when the tolerance is large, i.e., $\tau = 10^{-4}$. The amplitude of the oscillations is reduced when $\tau = 10^{-12}$. We expect the solution of the underlying differential algebraic equation to behave nicely, but we have no such expectation for the computed approximation unless $\tau$ is very small.

@@

Thirdly, if the computed energies for ${\tt tol} = 10 ^{-12}$ follow an asymptotic error expansion, then the commonly used time step of $1$ fs ($n=1000$ in this case) is \emph{not} well inside the asymptotic range. Why is this? If we were in the asymptotic range, then $\hat{A}_h \approx T - \alpha h^p$ would be a good approximation for some $\alpha \not = 0$ and $p>0$. In particular, the value of $\hat{A}_h$ should behave in a \emph{monotone} manner and the tiny oscillations that we have recorded should not be present.

@@

When the force-fields are not perturbed, see Figure~\ref{fig:iontrap_mwe1}, the experiments support the existence of an asymptotic error expansion and we can clearly identify an asymptotic range for each of the 4 Runge-Kutta methods used to integrate Newton's equations of motion.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{iontrap_mwe2.pdf}
  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p$. The fields are zero outside a small sphere and discontinuous.}
  \label{fig:iontrap_mwe2}
\end{figure}
When the force-fields are truncated and discontinuities are introduced into the simulation, see Figure~\ref{fig:iontrap_mwe2}, there is no evidence to support the existence of an asymptotic error expansion and there is no reason to suspect that Richardson's error estimate is accurate.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{iontrap_mwe4.pdf}
  \caption{The evolution of $F_h$ for the total kinetic energy of a system of ions using integrators of order $p$. The fields are zero outside a small sphere, but smooth.}
  \label{fig:iontrap_mwe4}
\end{figure}
When the force-fields vanish outside a small ball but are smooth, see Figure~\ref{fig:iontrap_mwe4}, we retain the ability to judge when Richardson's error estimates are reliable and we can quantify the effect of changing the force-field.
